NumPy---Array Computation for Python

### 2019-12-20 paper.tex
Update recent technical improvements and discussion sections

% Intel, QuanSight, ...

@@

The central position which NumPy takes within the numerical and scientific
Python ecosystem  means that a vast number of projects are built on it.
These projects are consumers of the NumPy API.

@@

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of Array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.
A brief example of such a function which is written using NumPy
but can now be used more generally is:
\begin{lstlisting}
def f(x):
    y = np.tensordot(x, x.T)
    return np.mean(np.exp(y))
\end{lstlisting}
For example, a user relying on this function and normally using it with a
NumPy array, may now use a distributed Dask array when challenged
by a very large dataset, without any modification to \code{f(x)}.

%In order to aid better interaction between NumPy and these libraries,
%an ``array function'' protocol was added to NumPy in 2018 (see Section
%X). The new protocol allows for external array objects (sometimes
%called ``duck arrays'') to pass through NumPy:
%
%\begin{lstlisting}
%  import numpy as np
%  import cupy as cp
%
%  x_gpu = cp.array([1, 2, 3])
%  y = np.sum(x_gpu)  # This works, and returns a GPU array
%\end{lstlisting}

@@

Most of the technical work for the Moore and Sloan grant is now complete.
The main remaining item is an overhaul of the data type system, which is discussed
below.  Now that years of technical debt have been addressed, the project
is in good shape for growing the community and better addressing the needs
of new users.  That is the main focus of the CZI grant, which we discuss
below.

@@

The meaning of each element within a NumPy array is described by its datatype
(dtype). NumPy arrays can hold all common numerical datatypes, string, datetimes
and generic Python objects, each of these is identified by a dtype.
We are busy overhauling the datatype system to make it consistent (aligned with
expectations), to simplify the creation of custom dtypes.
While numerical dtypes make the foundation for most users, there are
a multitude of use cases which cannot prosper due to current limitations.
These use cases include datatypes with physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, automatic differentiation\cite{pyadolc}, and many more.
Projects addressing these exist, each struggles with current
limitations.  % TODO: I may need citations, we could link github issues.

While it is currently possible to define custom dtypes within NumPy, many of the
use cases above cannot be solved using this system.
This is evidenced also by code within NumPy which would require changes
in many places to add such dtypes, even when adding them within NumPy itself.
The new implementation will be simpler and more cleanly implemented in
this regard.
Custom dtypes and the basic dtypes included within NumPy will be implemented
and behave identically as much as possible.
At the same time redesigning the API will make NumPy dtypes more flexible to grow
to future needs should they arise.

As a further step, to make dtypes more accessible and allow a faster adoption of
custom dtypes within the community, we plan to create a Python API to
define such custom dtypes.

A \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}
is under development in consultation with the community.

>>>>

NumPy has, throughout its lifetime, seen continuous development.
E.g., bugs fixes, minor improvements, cleanups, and releases have
happened regularly.  However, with the NEP process in place we now
have the flexibility to plan and implement changes of larger scope.
We highlight two of those below, as well as changes made to our
testing infrastructure to support hardware platforms used in large
scale computing.

@@

The central position of NumPy takes within the numerical and scientific
Python ecosystem means that a vast number of projects are built on it.
These projects are consumers of the NumPy API.

@@

To address these issues NumPy has the goal of providing the fundamental
API for \emph{interoperability} between the various NumPy-like APIs.
An earlier step in this direction was the implementation of the
\code{\_\_array\_ufunc\_\_} protocol in NumPy 1.13, which enabled interoperability
for most mathematical functions.\cite{NEP13}
In 2019 this was expanded more generally with the inclusion of the
\code{\_\_array\_function\_\_} protocol into NumPy~1.17.
These two protocols allows providers of Array-like APIs to be interoperable
with the NumPy API: their arrays work correctly with almost all NumPy functions.\cite{NEP18}
For the users relying on specialized array projects it means that even though
much code is written specifically for NumPy arrays and uses the NumPy API as
\code{import numpy as np}, it can nevertheless work for them.  For
example, here is how a CuPy GPU array can be passed through NumPy for
processing, with all operations being dispatched back to CuPy:

\begin{lstlisting}
  import numpy as np
  import cupy as cp

  x_gpu = cp.array([1, 2, 3])
  y = np.sum(x_gpu)  # This works, and returns a GPU array
\end{lstlisting}

Similarly, user defined functions composed using NumPy can now be
applied to, e.g., multi-node distributed Dask arrays:

\begin{lstlisting}
  import numpy as np
  import dask.array as da

  def f(x):
      """Function using NumPy API calls"""
      y = np.tensordot(x, x.T)
      return np.mean(np.log(y + 1))

  x_local = np.random.random([10000, 10000])  # random local array
  x_distr = da.random.random([10000, 10000])  # random distributed array

  f(x_local)  # returns a NumPy array
  f(x_distr)  # works, returns a Dask array
\end{lstlisting}

@@

Most of the technical work for the Moore and Sloan grant is now complete.
The main remaining item is an overhaul of the data type system.

The meaning of each element within a NumPy array is described by its
datatype (dtype). NumPy arrays can hold all common numerical
datatypes, strings, datetimes, and generic Python objects, each of
which is identified by a dtype.  We are busy overhauling the datatype
system to make it behave consistently and to simplify the creation of
custom dtypes, both in C and in Python [XXX Cite the NEP here, instead
of \href{https://github.com/numpy/numpy/pull/14422}{draft of DTypes NEP}].
While numerical dtypes make the foundation for most users,
there are a multitude of use cases which cannot prosper due to current
limitations, such as physical units\cite{astropy,Goldbaum2018,pint},
% pyadolc may be a bit too small a project, so may want to remove/replace with an other example.
geometrical objects\cite{pygeos}, and automatic
differentiation\cite{pyadolc}.
%Projects addressing these exist, each struggles with current
%limitations.  % TODO: I may need citations, we could link github issues.

%% While it is currently possible to define custom dtypes within NumPy, many of the
%% use cases above cannot be solved using this system.
%% This is evidenced also by code within NumPy which would require changes
%% in many places to add such dtypes, even when adding them within NumPy itself.
%% The new implementation will be simpler and more cleanly implemented in
%% this regard.
%% Custom dtypes and the basic dtypes included within NumPy will be implemented
%% and behave identically as much as possible.
%% At the same time redesigning the API will make NumPy dtypes more flexible to grow
%% to future needs should they arise.

%% As a further step, to make dtypes more accessible and allow a faster adoption of
%% custom dtypes within the community, we plan to create a Python API to
%% define such custom dtypes.

Now that we have made good progress on addressing technical debt, the project
is ready to grow its community of contributors and to better address the needs
of an ever-growing number of new users.  This is the main focus of the
forthcoming CZI grant.

[... CZI grant ...]

[... welcome new users ...]

@@
